Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-09 submission/manuscript.Rmd
update XY EpiLC to XX vs XY deseq2 results

We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. The majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells. Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). \textbf{B.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation. Brightfield images taken at 20X. \textbf{C.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs. Welch's t-test, expression in transcripts per million (TPM). \textbf{D.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs. * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets. \textbf{F.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue). \textbf{G.} enrichPlot gene ontology analysis comparing enriched biological processes for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs. \textbf{H.} Top left - Example germline identity DEGs unique to EpiLCs. Top right - Example 2-cell genes that are not dysregulated in \textit{Kdm5c}-KO EpiLCs. p-values for Welch's t-test. Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs.}
  \label{figurelabel}
\end{figure}

>>>>

We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. The majority of germline DEGs were unique to either EpiLCs or the brain, with only _D1Pas1_ and _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells. Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). \textbf{B.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation. Brightfield images taken at 20X. \textbf{C.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs. Welch's t-test, expression in transcripts per million (TPM). \textbf{D.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs. * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets. \textbf{F.} Average bigwigs of an example germline gene, \textit{D1Pas1}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue). \textbf{G.} enrichPlot gene ontology analysis comparing enriched biological processes for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs. \textbf{H.} Top left - Example germline identity DEGs unique to EpiLCs. Top right - Example 2-cell genes that are not dysregulated in \textit{Kdm5c}-KO EpiLCs. p-values for Welch's t-test. Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs.}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs. \textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs. Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only). \textbf{D.} Heatmap of germline DEGs shared between male and female mutants. Color is the average log 2 fold change from sex-matched wild-type. \textbf{E.} Number of genes within each cluster from D. Clusters with higher expression in females compared to males (XX-biased) highlighted in orange. \textbf{F.} Example average bigwigs of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants. \textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome. Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant. X chromosome highlighted in black. \textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes. \textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants. \textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants.}
  \label{figurelabel}
\end{figure}
